Information-theoretic measures for non-linear causality detection: \\ application to social media sentiment and cryptocurrency prices

### 2020-07-24 paper/submission.tex
Attempt to clarify terminology re: causality vs statistical/Granger causality v2

  It is a feature of the nonparametric estimation of entropy that the absolute scale of the transfer entropy measure has only limited meaning. To detect causality, a relative position must be considered. A simple technique proposed by Marschinski \& Kantz \cite{Marschinski2002} is the Effective Transfer Entropy (ETE), derived by subtracting from the observed transfer entropy an average transfer entropy figure calculated over independently-shuffled time series, which destroys the temporal order and hence any causality.

  We adopt a shuffling approach producing 100 null-hypothesis transfer entropy values from independently shuffled time-series over the same domain, containing no causality. By calculating the mean and standard deviation of the shuffled transfer entropy figures, we estimate the significance of a causal result as the distance between the result and the average shuffled result, standardising by the shuffled standard deviation:

@@

  In order to validate the autoregressive and information-theoretic approaches to detecting causality, we apply these to the calculation of transfer entropy for synthetic data generated by both linear and non-linear coupled time series, of increasing coupling strength. 
  The computations are performed over the relative returns of the synthetic signals generated as described in sections \ref{s.method}\ref{s.LinearSignal} and \ref{s.method}\ref{s.NonLinearSignal}. 
  It is important to note that many areas for which such statistical studies are important are constrained by limited available data. Therefore, for a technique of causal inference to be useful, it must perform even with small sample sizes. In Section \ref{s.results} we apply both classical and information-theoretic methodologies to financial data with sample sizes of the order of $10^4$, and so we first evaluate the performance using synthetic time series of $15,000$ observations, with known causal coupling, before confirming that the technique can detect the same causal coupling with as few as $500$ samples. The results of this evaluation for increasing signal strength, using sample sizes of $500$, $1000$, $2500$, $10,000$ and $100,000$, are presented in Table {\color{blue}2} in the appendix.

@@

  Figure \ref{fig:CLM_confirmation} shows the mean transfer entropy results for 15,000 synthetic data points. We observe that, for this system, the linear method is incapable of detecting the causal relationship; it finds some significant information transfer (at a near-zero level, which we interpret as the linear component), but fails to represent the expected exposure–response relationship and does not correctly identify the expected anticipatory signal in the reverse direction. The information-theoretic method, by contrast, produces results which better represent the increasing coupling strength relationship, and direction of causality in the system. The technique also detects causality from $Y$ to $X$, for large values of $\alpha$. This effect was also observed by Liang and Hahs \& Pethel, and corresponds to the anticipatory nature of the coupling. Specifically the coupling function $g(x)$ involves repeated application of the update function $f(x)$ from equation \ref{eq:f(x)}, from which it can be seen that $f(f(X_t))$ is equivalent to $X_{t+2}$ so, for large $\alpha$, $Y_t$ will contain increasing amounts of the future information of $X_t$. In fact, at large coupling strengths approaching $\alpha=1$, the observed transfer entropy from $X$ to $Y$ begins to decrease, as more information exists in $Y$ about its future evolution.

@@

  \begin{figure*}[!htb]
    \includegraphics[width=\linewidth]{images/confirming_gbm.pdf}
    \caption{Demonstration that both linear and non-linear transfer entropy methods detect causality for linearly coupled synthetic data. The plots are calculated from the mean over 10 realisations using 15,000 data points of the synthetic random walk process from equations \ref{eq:GBM} and \ref{eq:coupled_random_walk} with $\mathbb{E} [\mu]=0.1$ and $\mathbb{E} [\mu^{\prime}]=0.1$, each with standard deviations of $0.1$, and $\mathbb{E} [\sigma]=0.025$ and $\mathbb{E} [\sigma^{\prime}]=0.025$, each with standard deviations of $0.01$.
    Non-linear transfer entropy is calculated using a quantile histogram of 6 classes per dimension. The Z-score of each result is also plotted for both methods. We observe no significant transfer entropy in the non-causal direction $Y \rightarrow X$. For the direction $X \rightarrow Y$, the Z-score of the linear technique is greater by orders of magnitude than the information-theoretic technique. However, both techniques correctly detect the linear transfer entropy, presenting largely equivalent results.}
    \label{fig:GBM_confirmation}

@@

The information-theoretic investigation detected a significant non-linear causal relationship in BTC, LTC and XRP, over multiple timescales and in both directions of sentiment to price and price to sentiment. The effect was strongest and most consistent for LTC and XRP, and in both cases the net information transfer was in the direction of sentiment to price. ETH and BTC were observed to show significant net information transfer in the direction price to sentiment, although separate dynamics were observed in ETH which see a collpase in significance. Given the hypothesis that low barriers to entry and unsophisticated investor speculation are key drivers for price movements, and that BTC and ETH represent the most widely known and traded cryptocurrencies, the fact that causality is detected most clearly for these currencies corresponds to expectations.

>>>>

  It is a feature of the nonparametric estimation of entropy that the absolute scale of the transfer entropy measure has only limited meaning. To detect causality, a relative position must be considered. A simple technique proposed by Marschinski \& Kantz \cite{Marschinski2002} is the Effective Transfer Entropy (ETE), derived by subtracting from the observed transfer entropy an average transfer entropy figure calculated over independently-shuffled time series, which destroys the temporal order and hence {\color{blue} the time-dependence required for statistical causality}.

  We adopt a shuffling approach producing 100 null-hypothesis transfer entropy values from independently shuffled time-series over the same domain, containing no {\color{blue} causal relationships}. By calculating the mean and standard deviation of the shuffled transfer entropy figures, we estimate the significance of a causal result as the distance between the result and the average shuffled result, standardising by the shuffled standard deviation:

@@

  In order to validate the autoregressive and information-theoretic approaches to detecting {\color{blue}statistical} causality, we apply these to the calculation of transfer entropy for synthetic data generated by both linear and non-linear coupled time series, of increasing coupling strength. 
  The computations are performed over the relative returns of the synthetic signals generated as described in sections \ref{s.method}\ref{s.LinearSignal} and \ref{s.method}\ref{s.NonLinearSignal}. 
  It is important to note that many areas for which such statistical studies are important are constrained by limited available data. Therefore, for a technique of causal inference to be useful, it must perform even with small sample sizes. In Section \ref{s.results} we apply both classical and information-theoretic methodologies to financial data with sample sizes of the order of $10^4$, and so we first evaluate the performance using synthetic time series of $15,000$ observations, with known causal coupling, before confirming that the technique can detect the same causal coupling with as few as $500$ samples. The results of this evaluation for increasing signal strength, using sample sizes of $500$, $1000$, $2500$, $10,000$ and $100,000$, are presented in Table {\color{blue}2} in the appendix.

@@

  Figure \ref{fig:CLM_confirmation} shows the mean transfer entropy results for 15,000 synthetic data points. We observe that, for this system, the linear method is incapable of detecting the causal relationship; it finds some significant information transfer (at a near-zero level, which we interpret as the linear component), but fails to represent the expected exposure–response relationship and does not correctly identify the expected anticipatory signal in the reverse direction. The information-theoretic method, by contrast, produces results which better represent the increasing coupling strength relationship, and direction of {\color{blue}Granger} causality in the system. The technique also detects {\color{blue}Granger} causality from $Y$ to $X$, for large values of $\alpha$. This effect was also observed by Liang and Hahs \& Pethel, and corresponds to the anticipatory nature of the coupling. Specifically the coupling function $g(x)$ involves repeated application of the update function $f(x)$ from equation \ref{eq:f(x)}, from which it can be seen that $f(f(X_t))$ is equivalent to $X_{t+2}$ so, for large $\alpha$, $Y_t$ will contain increasing amounts of the future information of $X_t$. In fact, at large coupling strengths approaching $\alpha=1$, the observed transfer entropy from $X$ to $Y$ begins to decrease, as more information exists in $Y$ about its future evolution.

@@

  \begin{figure*}[!htb]
    \includegraphics[width=\linewidth]{images/confirming_gbm.pdf}
    \caption{Demonstration that both linear and non-linear transfer entropy methods detect {\color{blue}statistical} causality for linearly coupled synthetic data. The plots are calculated from the mean over 10 realisations using 15,000 data points of the synthetic random walk process from equations \ref{eq:GBM} and \ref{eq:coupled_random_walk} with $\mathbb{E} [\mu]=0.1$ and $\mathbb{E} [\mu^{\prime}]=0.1$, each with standard deviations of $0.1$, and $\mathbb{E} [\sigma]=0.025$ and $\mathbb{E} [\sigma^{\prime}]=0.025$, each with standard deviations of $0.01$.
    Non-linear transfer entropy is calculated using a quantile histogram of 6 classes per dimension. The Z-score of each result is also plotted for both methods. We observe no significant transfer entropy in the non-causal direction $Y \rightarrow X$. For the direction $X \rightarrow Y$, the Z-score of the linear technique is greater by orders of magnitude than the information-theoretic technique. However, both techniques correctly detect the linear transfer entropy, presenting largely equivalent results.}
    \label{fig:GBM_confirmation}

@@

The information-theoretic investigation detected a significant non-linear causal relationship in BTC, LTC and XRP, over multiple timescales and in both directions of sentiment to price and price to sentiment. The effect was strongest and most consistent for LTC and XRP, and in both cases the net information transfer was in the direction of sentiment to price. ETH and BTC were observed to show significant net information transfer in the direction price to sentiment, although separate dynamics were observed in ETH which see a collpase in significance. {\color{blue}<removed>}%Given the hypothesis that low barriers to entry and unsophisticated investor speculation are key drivers for price movements, and that BTC and ETH represent the most widely known and traded cryptocurrencies, the fact that causality is detected most clearly for these currencies corresponds to expectations.
